Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2023-12-20 doc/manuscript/translocation_pnas.tex

% Please add a significance statement to explain the relevance of your work
\significancestatement{Understanding how species persist despite accelerating global change is
critical for the conservation of biodiversity. One possibility is rapid
evolution that restores positive growth to declining populations
(``evolutionary rescue''). Individuals from these rescued populations
could also be used to reestablish extinct populations. We evaluated both
possibilities in a once-common frog species driven to the brink of
extinction by a disease that has devastated global amphibian
biodiversity and for which few options exist to reverse disease impacts.
We found that frogs have evolved increased resistance to infection,
facilitating population recovery. In addition, introduction of frogs
from these populations allowed reestablishment despite the ongoing
presence of disease. Such reintroductions could play an important role
in biodiversity conservation in our rapidly changing world.}

@@

Disease-induced population declines are decimating global biodiversity
\citep{daszak2000}, but strategies to recover affected species are
generally lacking. Here, we tested the possibility that populations of
resistant/tolerant individuals could be used to reestablish extirpated
populations. Although previously proposed
\citep{mendelson2019, scheele2021}, this strategy has rarely been used
in the wild \citep[see][ for a small-scale example]{joseph2018}. In
particular, although numerous studies have documented signatures of
selection in hosts following disease-induced declines
\citep[e.g.,][]{epstein2016, gignoux-wolfsohn2021}, we are unaware of
any that used individuals bearing the resulting adaptive alleles to
reestablish extirpated populations. The results of our study provide
evidence that (1) MYL frogs have experienced strong selection following
exposure to a recently-emerged lethal fungal disease and that some of
the genomic regions under selection contain genes related to disease
resistance/tolerance, and (2) reintroduction of individuals from these
resistant/tolerant populations allows the successful reestablishment of
viable frog populations despite ongoing disease.

Previous field studies in MYL frogs show that frog-Bd dynamics and frog
survival in the presence of Bd are fundamentally different between naive
and recovering populations. Following the arrival and establishment of
Bd in previously-naive populations, adult frogs develop high Bd loads
that lead to mass die-offs \citep{vredenburg2010}. In contrast, in
recovering populations adult frogs typically have low-to-moderate and
relatively constant Bd loads and mass die-offs are not observed
\citep[see also
Figure S5]{briggs2010, knapp2011}. The
differences in Bd load of frogs from naive and recovering populations
are also observed in controlled laboratory studies \citep[see Figure 4
in][]{knapp2016}. The observed reduced susceptibility of frogs from
recovering populations could in theory be due to several factors,
including natural selection for more resistant/tolerant genotypes and/or
acquired immunity, but until now evidence to evaluate the role of
evolution was lacking.

Results from our genomic analyses strongly suggest that natural
selection for adaptive alleles and resulting increased
resistance/tolerance is at least partially responsible for the reduced
susceptibility of frogs in recovering populations. We identified
specific alleles and genomic regions showing signatures of selection
between adjacent Bd-naive and recovering MYL frog populations,
consistent with strong selection following Bd exposure. Importantly,
some regions under selection are associated with cellular and
immunological mechanisms known to contribute to disease resistance,
including in amphibians \citep{zamudio2020}. For example, the MHC plays
an important role in immunity. In our study, we identified a region that
shows evidence of selection in recovering populations and contains eight
genes associated with either the MHC Class I or Class III regions. These
results corroborate numerous previous studies linking MHC genes to
amphibian resistance/tolerance against Bd
\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
the strongest indication of directional selection (as measured by
\(\pi_{diff}\)) contains the interferon-related gene GVINP1. Several
previous studies have found this gene to be differentially expressed
during Bd infection \citep[e.g.,][]{grogan2018, ellison2014} and in
populations differing in Bd susceptibility \citep{grogan2018}. This gene
is also strongly linked to disease in salmon, explaining a notable 20\%
of the resistance phenotype \citep{robledo2020, robledo2018}. We also
identified a region, characterized by high \emph{F\textsubscript{ST}}
and low \(\pi_{diff}\), that contained the complement genes C6 and C7.
The complement system plays an important role in innate immunity
\citep{rieraromo2016}, and our results could indicate that balancing
selection is acting in this region of the genome to favor a diverse set
of alleles, as is known for C6 in humans \citep{soejima2005}. Finally,
based on the analysis of individual outlier variants, the RIN3 gene
showed a consistent pattern of allele frequency differences across all
nine of the frog populations sampled in this study, indicating
consistent selection in populations distributed across a wide geographic
area. This gene is associated with inflammation response and in
\emph{Xenopus} is highly expressed in wound stumps following amputation
of tadpole tails \citep{fukazawa2009}. Also in this analysis, the
outlier variant with the lowest p-value (uncharacterized gene
LOC108802036) is located adjacent to a type I interferon gene (Np-IFNi2)
in another frog genome \citep{gan2018}, and together with GVINP1 further
suggests the importance of interferon-related genes in this system.

Reintroduction of resistant/tolerant MYL frogs was remarkably successful
in reestablishing viable populations in the presence of Bd. Of the 12
translocated populations, approximately 80\% showed evidence of both
successful reproduction and recruitment of new adults. Bd loads measured
before and after translocation were similar and generally remained well
below the level at which frog mortality occurs. Annual survival for 12
of the 24 translocated cohorts exceeded 50\%, and \textgreater{} 70\% of
translocated cohorts had survival above this 50\% level when the
earliest translocations are excluded (i.e., translocations conducted
when methods were still being refined). The fact that the relatively low
Bd loads and correspondingly high frog survival in the presence of Bd
was maintained between the natal/donor sites and the recipient sites
indicates that these characteristics of naturally-recovering populations
were not solely an effect of the natal/donor site or site * frog
interactions, but instead were due at least in part to
resistance/tolerance inherent in the frogs. In addition, the relatively
high survival of translocated frogs was maintained in their progeny, as
expected if resistance/tolerance has a genetic basis. Results from the
population viability model were also encouraging. In particular,
translocated populations with \textgreater{} 50\% survival in the first
year post-translocation were predicted to have a low probability of
extinction over 50 years (probability of extinction = 0 - 0.32). In
light of the generally low success rate of amphibian reintroduction
efforts \citep{dodd2005}, our success in reestablishing MYL frog
populations via translocation of resistant/tolerant individuals is
striking, and even more so given that this success occurred despite
ongoing infection of frogs with Bd.

Despite the resistance/tolerance of adult MYL frogs in recovering
populations, individual and population-level impacts of Bd are still
evident. In an earlier study of 2 of our 12 translocated populations
\citep{joseph2018}, Bd infection and load had detectable effects on the
survival of adults and may have influenced population establishment
(sites referred to as ``Alpine'' and ``Subalpine'' in \citep{joseph2018}
are identified as ``70550'' and ``70505'' in the current study).
Applying similar analyses to all 12 of our translocated populations
would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on
adults, the impacts on younger life stages are more apparent. MYL frogs
immediately following metamorphosis (``metamorphs'') are highly
susceptible to Bd infection \citep{ellison2018} and as a result
experience high mortality. This high susceptibility of metamorphs is
documented in numerous species of anurans, and may result from the
poorly developed immune system characteristic of this life stage
\citep{humphries2022}. In naturally recovering and translocated MYL frog
populations, we suggest that the high mortality of metamorphs is an
important limitation on subsequent recruitment of new adults. Therefore,
although adult MYL frogs appear relatively resistant/tolerant, Bd
infection continues to have important limiting effects on recovering
populations \citep[see also][]{hollanders2022}.

The recent emergence of Bd worldwide has contributed to the decline of
hundreds of amphibian species, some of which are now extinct in the wild
\citep{scheele2019}. This unprecedented impact on global amphibian
biodiversity is compounded by the lack of any effective and broadly
applicable strategies to reverse these impacts
\citep{garner2016, knapp2022}. Importantly, in addition to the natural
recovery documented for MYL frogs \citep{knapp2016}, other species are
also showing evidence of post-epizootic recovery in the presence of Bd
\citep{scheele2017, voyles2018} and suggest the possibility of also
using animals from these recovering populations to reestablish
extirpated populations. As with MYL frogs, the feasibility and long-term
success of such efforts will depend on the availability of robust donor
populations containing individuals that have acquired via evolution the
adaptive alleles necessary to allow frog survival and population growth
in the presence of Bd. Despite the hopeful example of successful
reestablishment of MYL frogs via evolutionary rescue and
reintroductions, the challenge of recovering hundreds of Bd-impacted
amphibian species worldwide is a daunting prospect. Although we now have
a proven strategy to reestablish extirpated MYL frog populations,
recovery across their large historical range will require substantial
resources over many decades. The results of this study provide a hopeful
starting point for that endeavor and other future efforts worldwide.

Evolutionary rescue may be increasingly critical for the survival of
wildlife populations in a rapidly changing world. The importance of
evolutionary rescue is known for a wide variety of taxa
\citep[e.g.,][]{carlson2014, mills2018, oziolor2019}, including in the
specific context of disease-impacted systems \citep{searle2020}.
However, obtaining evidence in wild animal populations for all the
causal links in this process (i.e., selection for adaptive alleles
\(\rightarrow\) increase in host fitness in the presence of the stressor
\(\rightarrow\) positive population growth and recovery) is difficult in
most systems. Nonetheless, we provide a compelling example from the wild
not only that evolutionary rescue can produce individuals that harbor
adaptive alleles and allow population recovery, but importantly that
these individuals can be used in translocations to reestablish
extirpated populations. We expect that such evolutionary rescue-based
translocations will be an essential tool in wildlife conservation in an
era of accelerating global change.

\matmethods{
\hypertarget{frog-evolution-in-response-to-bd-1}{%
\section*{Frog evolution in response to
Bd}\label{frog-evolution-in-response-to-bd-1}}

\hypertarget{sampling-and-sequencing}{%
\subsection*{Sampling and Sequencing}\label{sampling-and-sequencing}}

We collected DNA samples via buccal swabbing \citep{broquet2007} from 53
\emph{Rana muscosa}/\emph{Rana sierrae} individuals: 24 from 4 naive
populations, and 29 from 5 recovering populations. These populations are
located in the southern Sierra Nevada, from northern Yosemite National
Park to northern Sequoia National Park
(Figure~\ref{fig-allelefrequencies}). Samples were collected from 5-6
frogs per population. To minimize potential confounding effects caused
by variation in frog genotypes across latitude
\citep[e.g.,][]{rothstein2020}, we selected sampling sites such that
both population types were represented across similar latitudinal
ranges. DNA was extracted following Qiagen DNEasy manufacturer's
protocols. We sequenced the samples using an exome capture assay as
described in Byrne et al.~(in review). Briefly, genomic libraries were
prepared and captured using a custom Nimblegen capture pool. Capture
baits were designed based on the coding regions of the \emph{R. muscosa}
transcriptome (GenBank accession GKCT00000000). Captured libraries were
then pooled and sequenced on a NovaSeq 6000 150PE Flow Cell S1 at the
Vincent J. Coates Genomics Sequencing Lab at UC Berkeley. Raw sequencing
reads are available from NCBI SRA (PRJNA870451).

\hypertarget{data-pre-processing-and-cleaning}{%
\subsection*{Data Pre-processing and
Cleaning}\label{data-pre-processing-and-cleaning}}

We then further filtered our dataset at the individual and variant
level. First, we trimmed our variants to only include those with minor
allele frequency \textgreater{} 0.03, a maximum depth of 250 and minimum
depth of 5, a minimum genotype quality of 20, and a maximum missing
proportion of 0.5. This filter resulted in 427,038 sites, of which
353,172 were SNPs and 73,866 were INDELS. Finally, we trimmed samples
with an average depth across filtered sites \textless{} 7x. Our final
dataset included 50 samples, 23 from naive and 27 from recovering
populations, with an average depth of 16.7x (range = 7.4x -- 26.1x).

\hypertarget{data-analysis}{%
\subsection*{Data Analysis}\label{data-analysis}}

To visualize the genomic relationships of our populations we conducted a
PCA using the glPCA function in the adegenet R package
\citep{jombart2008}. To detect regions of the genome under selection
between naive and recovering populations, we used two approaches: (1) a
multivariate linear mixed model to evaluate individual variants (SNPs
and INDELs), and (2) a splined window analysis to evaluate larger
genomic regions. For the variant analysis, we first used a stringent
data filter to include only variants with \textless{} 5\% missing data
(missing for no more than 2 individuals), and then calculated the
likelihood ratio statistic for the resulting set of 148,307 high quality
variants across 127 contigs using GEMMA \citep{zhou2014}. GEMMA
calculates and incorporates a relatedness matrix for input samples,
allowing us to account for relatedness and population structure when
calculating likelihood ratio statistics. We identified variants showing
different allele frequencies between naive versus recovering populations
(``outliers'') using a Bonferroni-corrected significance level of 0.01.
We visualized the results using a Manhattan plot and qqplot. We
developed a more liberal set of outlier variants using a
Bonferroni-corrected significance level of 0.05 and used this set solely
for the gene ontology (GO) analysis (see below and SI Appendix: Dataset
S1, S2).

In the splined window analysis, we identified outlier regions using
\emph{F\textsubscript{ST}} and differences in nucleotide diversity
(\(\pi_{diff}\)) between naive and recovering populations. First, we
calculated per-site \emph{F\textsubscript{ST}} between the naive and
recovering individuals for all bi-allelic SNPs in the 30 largest contigs
(98\% of all SNPs) using VCFtools \citep{danecek2011}. Next, we
calculated per-site nucleotide diversity \(\pi\) separately for
individuals from the naive and recovering populations using VCFtools,
then calculated \(\pi_{diff}\) for each population
(\(\pi_{diff} = \pi_{naive} - \pi_{recovering}\)). We concatenated the
values for \emph{F\textsubscript{ST}} and \(\pi_{diff}\) in order of
size-sorted chromosome number and adjusted the SNP position based on the
relative position in the genome. We then used the GenWin R package
\citep{beissinger2015} to conduct a splined discrete window analysis for
\emph{F\textsubscript{ST}} and \(\pi_{diff}\). This method calculates
where non-overlapping window boundaries should occur by identifying
inflection points in the fitted spline, therefore balancing false
positive and false negative results that occur using other window-based
methods \citep{beissinger2015}. This method also calculates a
W-statistic allowing for outlier identification. We identified outliers
as those with a W-statistic greater than 4 standard deviations above the
mean for \emph{F\textsubscript{ST}} or above/below the mean for
\(\pi_{diff}\). Shared outliers were then identified as those that were
outliers in both analyses, meaning that they showed (1) high
differentiation between naive and recovering populations, and (2)
differential patterns of nucleotide diversity in the same region.
Finally, we extracted gene transcripts mapped within each region and
retrieved annotation for that region using BLAST (SI Appendix: Datasets
S3, S4).

Using the liberal set of outlier variants, we performed a statistical
overrepresentation test (Fisher's exact test) using Blast2GO
\citep{conesa2005}. First, we retrieved the BLAST hits and mapped GO
terms for each gene in our targeted transcriptome. Next, we compared the
30 unique outlier genes from the liberal set of outlier variants to the
complete set of genes in our target transcriptome to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. We then repeated this process for the set of 33 genes
located in the overlapping \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows. Code used for genomic analyses and to create figures is
available from the following GitHub repository:
\url{https://github.com/allie128/mylf-selection}.

\hypertarget{frog-population-recovery-1}{%
\section*{Frog population recovery}\label{frog-population-recovery-1}}

@@

For the 24 translocations we conducted, we identified donor populations
from which adult frogs (\(\geq\) 40 mm snout-vent length) could be
collected using several years of visual encounter surveys (VES) and skin
swab collections \citep{knapp2016}, and results from population genetic
analyses \citep{poorten2017}. The populations that we selected contained
several hundred \emph{R. sierrae} adults and hundreds or thousands of
tadpoles. These relatively high abundances were the result of recent
increases following previous Bd-caused declines \citep{knapp2016}. As is
typical for recovering MYL frog populations, Bd prevalence in the donor
populations was high (0.69--0.96) and Bd load (median
log\textsubscript{10}(load) = 3.06--3.78 ITS copies) was two or more
orders of magnitude below the level at which frog mortality is expected
(log\textsubscript{10}(load) \(\approx\) 5.78
copies)\citep{vredenburg2010, joseph2018}. Recipient sites to which
frogs were translocated were chosen based on previous \emph{R. sierrae}
presence (determined from VES and/or museum records) or characteristics
that suggested high quality habitat for this species \citep{knapp2005}.

We conducted 1--4 translocations per site (Figure S3,
Figure~\ref{fig-translocation-survival}) and each translocated cohort
included 18 to 99 frogs (median = 30). In preparation for each
translocation, adult frogs were collected from the donor population and
measured, weighed, swabbed, and PIT tagged. Frogs were transported to
the recipient site either on foot or via helicopter. Following release,
translocated populations were visited approximately once per month
during the summer active season (median number of visits per summer = 2;
range = 1--10) and assessed using diurnal CMR surveys and VES. CMR
surveys allowed estimation of adult survival, adult population size, and
recruitment of new adults, and VES provided estimates of tadpole and
juvenile abundance. During 2006-2012, CMR surveys were conducted on a
single day (primary period) per site visit, during which all habitats
were searched repeatedly for adult frogs. Frogs were captured using
handheld nets, identified via their PIT tag (or tagged if they were
untagged), measured, weighed, swabbed, and released at the capture
location. During 2013-2022, we generally used a robust design in which
all habitats were searched during several consecutive days (median
number of secondary periods per primary period = 3; range = 3--7), and
frogs were processed as described above. However, when the number of
frogs detected on the first survey day was zero or near zero, we
typically conducted only a single-day CMR survey. When using a robust
design, within a primary period, frogs that were captured during more
than one secondary period were measured, weighed, and swabbed during the
first capture, and during subsequent captures were only identified and
released.

@@

For each of the translocated populations, we used the presence of
tadpoles and/or juveniles from VES and counts of new recruits (i.e,
untagged adults) in CMR surveys to provide two measures of successful
reproduction. To calculate the proportion of years in which
tadpoles/juveniles were present at a site, we excluded surveys conducted
in the year of the initial translocation to that site. This exclusion
accounted for the fact that all translocations were conducted after the
breeding period and reproduction would therefore not have occurred until
the following year. Similarly, to calculcate the proportion of years in
which new recruits were present at a site, we excluded surveys conducted
during the 3 years following the initial translocation. This accounted
for the multi-year tadpole and juvenile stages in MYL frogs
(Table S2).

@@

We estimated survival of translocated frogs, recruitment of new frogs
into the adult population, and adult population size using a Bayesian
open-population Jolly-Seber mark-recapture model with known additions to
the population (i.e., translocated cohorts), as implemented by the mrmr
package \citep{joseph2019} and using R Statistical Software
\citep[v4.4.4,][]{rsoftware2022}. The model tracks the states of
\emph{M} individuals that comprise a superpopulation made up of real and
pseudo-individuals \citep[see][ for details]{joseph2018}. The possible
states of individuals include ``not recruited'', ``alive'', and
``dead''. The possible observations of individuals include ``detected''
and ``not detected''. We assume that individuals that are in the ``not
recruited'' or ``dead'' states are never detected (i.e., there are no
mistakes in the individual PIT tag records). For all models, we used
mrmr defaults for priors, number of chains (4), and warmup and
post-warmup iterations (2000 for each). We evaluated convergence of the
Markov chain Monte Carlo (MCMC) algorithm using trace plots and
Gelman-Rubin statistics (Rhat).

@@

\clearpage

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Understanding how species persist despite accelerating global change is
critical for the conservation of biodiversity. Emerging infectious
diseases can have particularly devastating impacts, and few options
typically exist to reverse these effects. We used large-scale
reintroductions of disease-resistant individuals in an effort to recover
a once-common frog species driven to the brink of extinction by a
disease that has recently devastated global amphibian biodiversity.
Introduction of resistant frogs allowed reestablishment of populations
in the presence of disease. In addition, resistance appears to have a
genetic basis and is consistent with evolution following exposure of
frogs to Bd. The evolution of resistance and reintroductions of
resistant individuals could play an important role in biodiversity
conservation in our rapidly changing world.}

@@

Disease-induced population declines are decimating global biodiversity
\citep{daszak2000}, but broadly-applicable strategies to recover
affected species are generally lacking \citep[e.g.,][]{garner2016}. Our
results indicate (i) the capacity of reintroduced populations to become
established and eventually recover despite ongoing disease, (ii) that
the recovering populations are likely to persist over a 50-year period,
(iii) that there are substantial genomic differences between naive and
recovering MYL frog populations, consistent with evolutionary change in
frogs following Bd exposure, and (iv) that some of the genomic regions
putatively under selection contain genes related to disease resistance.
Collectively, these results (Figure~\ref{fig-recovery-model}) provide a
rare example of amphibian recovery in the presence of a highly virulent
pathogen, and have important implications for the conservation and
recovery of amphibians and other taxa worldwide that are endangered by
escalating impacts from emerging infectious diseases. In light of the
generally low success rate of amphibian reintroduction efforts
\citep{dodd2005}, our success in reestablishing MYL frog populations via
translocation of resistant individuals is striking, and even more so
given that this success occurred despite ongoing infection of frogs with
Bd.

Previous field studies in MYL frogs show that frog-Bd dynamics and frog
survival in the presence of Bd are fundamentally different between naive
and recovering populations. Following the arrival and establishment of
Bd in previously-naive populations, adult frogs develop high Bd loads
that lead to mass die-offs \citep{vredenburg2010}. In contrast, in
recovering populations adult frogs typically have low-to-moderate and
relatively constant Bd loads and mass die-offs are not observed
\citep[see also Figure S2]{briggs2010, knapp2011}. The differences in Bd load of frogs from
naive and recovering populations are also observed in controlled
laboratory studies \citep[see Figure 4 in][]{knapp2016}, and clearly
indicate that frogs from recovering populations exhibit resistance
against Bd infection. This resistance could in theory be due to several
factors, including natural selection for more resistant genotypes,
acquired immunity, and/or inherent between-population differences that
pre-date Bd exposure, but until now evidence to evaluate the role of
evolution was lacking.

Results from our genomic analyses suggest that natural selection for
adaptive alleles is at least partially responsible for the increased
resistance of frogs in recovering populations. We identified specific
alleles and genomic regions showing signatures of selection between
adjacent naive and recovering MYL frog populations, consistent with
selection following Bd exposure. Importantly, some regions under
selection are associated with cellular and immunological mechanisms
known to contribute to disease resistance, including in amphibians
\citep{zamudio2020}. For example, the MHC plays an important role in
immunity. In our study, we identified a region that shows evidence of
selection in recovering populations and contains eight genes associated
with either the MHC Class I or Class III regions. These results
corroborate numerous previous studies linking MHC genes to amphibian
resistance against Bd \citep[e.g.,][]{savage2011, bataille2015}.
Similarly, the region with the strongest indication of directional
selection (as measured by \(\pi_{diff}\)) contains the
interferon-related gene GVINP1. Several previous studies of amphibians
have found this gene to be differentially expressed during Bd infection
\citep[e.g.,][]{grogan2018b, ellison2014} and in populations differing
in Bd susceptibility \citep{grogan2018b}. This gene is also strongly
linked to disease in salmon, explaining a notable 20\% of the resistance
phenotype \citep{robledo2020, robledo2018}. We also identified a region,
characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
that contained the complement genes C6 and C7. The complement system
plays an important role in innate immunity \citep{rieraromo2016}, and
our results could indicate that balancing selection is acting in this
region of the genome to favor a diverse set of alleles, as is known for
C6 in humans \citep{soejima2005}. Finally, based on the analysis of
individual outlier variants, the RIN3 gene showed a consistent pattern
of allele frequency differences across all nine of the frog populations
sampled in this study, indicating consistent selection in populations
distributed across a wide geographic area. This gene is associated with
inflammation response and in \emph{Xenopus} is highly expressed in
wounds \citep{fukazawa2009}. In addition, the outlier variant with the
lowest p-value was the uncharacterized gene LOC108802036. In the genome
of another frog species, this gene is located adjacent to a type I
interferon gene (Np-IFNi2) \citep{gan2018}, and together with GVINP1
further suggests the importance of interferon-related genes in this
system. Collectively, the genes associated with these genomic
differences may confer at least some degree of resistance against Bd
infection, an attribute that may be critically important to population
reestablishment and recovery in the presence of Bd.

Reintroduction of resistant MYL frogs was remarkably successful in
reestablishing viable populations in the presence of Bd. Of the 12
translocated populations, approximately 80\% showed evidence of both
successful reproduction and recruitment of new adults. Year-1 survival
for 12 of the 24 translocated cohorts exceeded 50\%, and \textgreater{}
70\% of translocated cohorts had survival above this 50\% level when the
earliest translocations are excluded (i.e., translocations conducted
when methods were still being refined; see \textbf{Materials and Methods
- Frog population recovery - Field methods} for a brief description of
these refinements). The fact that the relatively low Bd loads and
correspondingly high frog survival was maintained when frogs were moved
from donor populations to recipient sites indicates that these
characteristics of naturally-recovering populations were not solely an
effect of site characteristics, but were also strongly influenced by
resistance inherent in the frogs. Although it could be argued that the
relatively invariant Bd loads before versus after translocation are a
consequence of similar pathogen pressure in the donor and translocated
populations, this is at odds with the fact that in the first year after
translocation frog densities are typically 1-2 orders of magnitude lower
in the translocated versus donor populations and pathogen pressure
should follow a similar pattern. In addition to the maintenance of Bd
load and frog survival between natal and translocation sites, the
relatively high survival of translocated frogs was maintained in their
progeny, as expected if resistance has a genetic basis.

Results from the population viability model were also encouraging. In
particular, translocated populations with \textgreater{} 50\% survival
in the first year post-translocation were predicted to have a low
probability of extinction over 50 years (probability of extinction \(<\)
0.5 when year 1 juvenile survival probability was greater than 0.10).
The viability model highlighted the important role of frog survival in
affecting long-term population viability, and allowed us to extend the
temporal scale of our study beyond the years covered by our
post-translocation surveys. These long-term forecasts are important,
given that reintroduced MYL frog populations may often take decades to
achieve our ultimate goal of self-sustainability \citep{joseph2018}.
Making well-supported projections about the long-term outcome of
reintroduction efforts from shorter-term information is critically
important to the process of adaptive management of species
reintroduction programs \citep{seddon2007}, including the one we are
carrying out for MYL frogs. Specifically, the combined results from our
reintroduction study and viability modeling indicate that survival of
frogs in the first year following translocation is an effective proxy of
longer-term survival and population viability. In addition, given the
repeatability of frog survival at a site, 1-year frog survival also
serves as an effective proxy of site quality (i.e., the ability of a
site to support high frog survival and a viable frog population over the
long term). This proxy of site quality is important in the MYL frog
system because accurately predicting the ability of a site to support a
viable frog population \emph{a priori} remains difficult, even after
conducting 24 translocations over 16 years.

Despite the demonstrated resistance of adult MYL frogs against Bd
infection, individual and population-level impacts of Bd are still
evident. In an earlier study of 2 of our 12 translocated populations
\citep{joseph2018}, Bd infection and load had detectable effects on the
survival of adults and may have influenced population establishment
(sites referred to as ``Alpine'' and ``Subalpine'' in \citep{joseph2018}
are identified as ``70550'' and ``70505'' in the current study).
Applying similar analyses to all 12 of our translocated populations
would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on
adults, the impacts on younger life stages are more apparent. MYL frogs
immediately following metamorphosis (``metamorphs'') are highly
susceptible to Bd infection \citep{ellison2019} and as a result
experience high mortality \citep{rachowicz2006}. This high
susceptibility of metamorphs is documented in numerous species of
anurans, and may result from the poorly developed immune system
characteristic of this life stage \citep{humphries2022}. In naturally
recovering and translocated MYL frog populations, we suggest that the
high mortality of metamorphs is an important limitation on subsequent
recruitment of new adults. Therefore, although adult MYL frogs appear
relatively resistant, Bd infection continues to have important limiting
effects on recovering populations \citep[see also][]{hollanders2022}.

The recent emergence of Bd worldwide has contributed to the decline of
hundreds of amphibian species, some of which are now extinct in the wild
\citep{scheele2019}. This extraordinary impact on global amphibian
biodiversity is compounded by the lack of any effective and broadly
applicable strategies to reverse these impacts
\citep{garner2016, knapp2022}. Importantly, in addition to the natural
recovery documented for MYL frogs \citep{knapp2016}, other species are
also showing evidence of post-epizootic recovery in the presence of Bd
\citep{scheele2017, voyles2018} and suggest the possibility of also
using animals from these recovering populations to reestablish
extirpated populations. As with MYL frogs, the feasibility and long-term
success of such efforts will depend on the availability of robust donor
populations containing individuals that have the adaptive alleles
necessary to allow frog survival and population growth in the presence
of Bd. Despite the hopeful example of successful reestablishment of MYL
frogs despite ongoing Bd infection, the challenge of recovering hundreds
of Bd-impacted amphibian species globally is a daunting prospect.
Although we now have a proven strategy to reestablish extirpated MYL
frog populations, recovery across their large historical range will
require substantial resources over many decades. The results of this
study provide a hopeful starting point for that endeavor and other
future efforts worldwide.

In our rapidly changing world, evolution is likely to play in important
role in facilitating the resilience of wildlife populations. Whether the
documented disease resistance in MYL frogs and concurrent recovery of
decimated populations provides an example of evolutionary rescue will
likely always be uncertain. Regardless, we provide an example from the
wild that suggests that evolution can produce individuals that harbor
adaptive alleles and allow population recovery in a novel (i.e.,
Bd-positive) environment, and show conclusively that individuals from
these recovering populations can be used to reestablish extirpated
populations and expand the scale of natural recovery
(Figure~\ref{fig-recovery-model}). We expect that similar species
recovery actions will be an essential tool in wildlife conservation in
an era of accelerating global change.

\matmethods{
\hypertarget{frog-population-recovery-1}{%
\section*{Frog population recovery}\label{frog-population-recovery-1}}

@@

For the 24 translocations we conducted, we identified donor populations
from which adult frogs (\(\geq\) 40 mm snout-vent length) could be
collected using several years of VES and skin swab collections
\citep{knapp2016}, and results from population genetic analyses
\citep{poorten2017}. The populations that we selected contained hundreds
to thousands of both \emph{R. sierrae} adults and tadpoles. These
relatively high abundances were the result of recent increases following
previous Bd-caused declines \citep{knapp2016}. As is typical for
recovering MYL frog populations, Bd prevalence in the donor populations
was high (0.69--0.96) and Bd load (median log\textsubscript{10}(load) =
3.06--3.78 ITS copies) was two or more orders of magnitude below the
level at which frog mortality is expected (log\textsubscript{10}(load)
\(\approx\) 5.78 copies)\citep{vredenburg2010, joseph2018}. Recipient
sites to which frogs were translocated were chosen based on previous
\emph{R. sierrae} presence (determined from VES and/or museum records)
or characteristics that suggested high quality habitat for this species
\citep{knapp2005}. At the beginning of this study, we had a relatively
limited understanding of the factors that affect habitat quality. In
subsequent years, we improved our site selection process by
incorporating new information about important habitat features, in
particular, overwinter habitats such as submerged boulders and
overhanging banks. \emph{R. sierrae} were absent from all recipient
sites prior to the first translocation.

We conducted 1--4 translocations per site
(Figure~\ref{fig-translocation-survival}, Figure S1)
and each translocated cohort included 18 to 99 frogs (median = 30). In
preparation for each translocation, adult frogs were collected from the
donor population and measured, weighed, swabbed, and PIT tagged. Frogs
were transported to the recipient site either on foot or via helicopter.
Following release, we visited translocated populations approximately
once per month during the summer active season and conducted diurnal CMR
surveys and VES (summer active season is generally July-August but can
start as early as May and end as late as September; range of survey
dates = May-25 to Sep-29, range of translocation dates = Jun-28 to
Sep-02; median number of visits per summer = 2, range = 1--10). CMR
surveys allowed estimation of adult survival, recruitment of new adults,
and adult population size, and VES provided estimates of tadpole and
juvenile abundance. During 2006-2012, we conducted CMR surveys on a
single day (primary period) per site visit, during which we searched all
habitats repeatedly for adult frogs. Frogs were captured using handheld
nets, identified via their PIT tag (or tagged if they were untagged),
measured, weighed, swabbed, and released at the capture location. During
2013-2022, we generally used a robust design in which all habitats were
searched during several consecutive days (median number of secondary
periods per primary period = 3; range = 3--7), and frogs were processed
as described above. However, when the number of frogs detected on the
first survey day was zero or near zero, we typically conducted only a
single-day CMR survey. When using a robust design, within a primary
period, frogs that were captured during more than one secondary period
were measured, weighed, and swabbed during the first capture, and during
subsequent captures were only identified and released.

@@

For each of the translocated populations, we used the presence of
tadpoles and/or juveniles from VES and counts of new recruits (i.e,
untagged adults) in CMR surveys to provide two measures of successful
reproduction. To calculate the proportion of years in which
tadpoles/juveniles were present at a site, we excluded surveys conducted
in the year of the initial translocation to that site. This exclusion
accounted for the fact that all translocations were conducted after the
breeding period and reproduction would therefore not occur until the
following year. Similarly, to calculate the proportion of years in which
new recruits were present at a site, we excluded surveys conducted
during the 3 years following the initial translocation. This accounted
for the multi-year tadpole and juvenile stages in MYL frogs
(Table S2 SI).

@@

For each translocation site, we estimated survival of translocated
frogs, recruitment of new frogs into the adult population, and adult
population size using a site-specific Bayesian open-population
Jolly-Seber CMR model with known additions to the population (i.e.,
translocated cohorts), as implemented by the mrmr package
\citep{joseph2019} and using R Statistical Software
\citep[v4.4.4,][]{rsoftware2022} (see \textbf{Supporting Information -
Frog population recovery - CMR model structure} for details). Briefly,
the model tracks the states of \emph{M} individuals that comprise a
superpopulation made up of real and pseudo-individuals \citep[see][ for
details]{joseph2018}. The possible states of individuals include ``not
recruited'', ``alive'', and ``dead''. The possible observations of
individuals include ``detected'' and ``not detected''. We assume that
individuals that are in the ``not recruited'' or ``dead'' states are
never detected (i.e., there are no mistakes in the individual PIT tag
records). We also assume that new recruits were the result of
within-site reproduction and not immigration from adjacent populations.
This assumption is justified by the fact that no \emph{R. sierrae}
populations were present within several kilometers of the translocation
sites. For all models, we used mrmr defaults for priors, number of
chains (4), and warmup and post-warmup iterations (2000 for each). We
evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm
using trace plots and Gelman-Rubin statistics (Rhat).

@@
